### 2025-07-25 paper/quarto_aaai/paper.tex
ready to submit

\subcaption{\label{fig-mnist-ce}\emph{Plausibility}: faithful
counterfactuals (\emph{ECCCo}).}

@@

\caption{\label{fig-mnist}Visual explanations for \emph{MNIST}. Top and
bottom rows (of images) show the results for BL and CT, respectively.
Mutability constraints are imposed on the five top and bottom rows (of
pixels). CT produces plausible counterfactuals (a) and is insensitive to
protected features (b).}

@@

For synthetic datasets, we always protect the first feature; for all
real-world tabular datasets we could identify and protect an \emph{age}
variable; for \emph{MNIST}, we protect the five top and bottom pixel
rows of the full image. Statistically significant reductions in costs
overwhelmingly point in the positive direction reaching up to around
60\% for \emph{GMSC}. Only in the case of \emph{Cred}, average costs
increase, likely because any benefits from protecting \emph{age} are
outweighed by an increase in costs required for greater plausibility.
The findings for \emph{Adult} and \emph{MNIST} are insignificant. A
qualitative inspection of the class-conditional integrated gradients in
Figure~\ref{fig-mnist-ig} suggests that CT still has the expected
effect: the model (bottom) is insensitive (blue) to the protected rows
of pixels; details are reported in the appendix.

>>>>

\subcaption{\label{fig-mnist-ce}\emph{Plausibility}: \emph{ECCCo}
counterfactuals for factual `0' (blue).}

@@

\caption{\label{fig-mnist}Visual explanations for \emph{MNIST}. Top and
bottom rows (of images) show the results for BL and CT, respectively.
Mutability constraints are imposed on the five top and bottom rows (of
pixels). CT produces plausible counterfactuals (a) and is less sensitive
to protected features (b).}

@@

For synthetic datasets, we always protect the first feature; for all
real-world tabular datasets we could identify and protect an \emph{age}
variable; for \emph{MNIST}, we protect the five top and bottom pixel
rows of the full image. Statistically significant reductions in costs
overwhelmingly point in the positive direction reaching up to around
60\% for \emph{GMSC}. Only in the case of \emph{Cred}, average costs
increase, likely because any benefits from protecting \emph{age} are
outweighed by an increase in costs required for greater plausibility.
The findings for \emph{Adult} and \emph{MNIST} are insignificant. A
qualitative inspection of the class-conditional integrated gradients in
Figure~\ref{fig-mnist-ig} suggests that CT still has the expected
effect: the model (bottom) is less sensitive (blue) to the protected
rows of pixels; details are reported in the appendix.

### 2025-07-25 paper/sections/other/main_results.qmd
ready to submit

To calculate the contributions, IG compares the output to a baseline. The selection of an appropriate baseline is an important design decision — it should produce a "neutral" prediction to avoid capturing effects that cannot be directly attributed to the model [@sundararajan2017ig; @sturmfels2020visualizing]. To remain consistent in our evaluations, we use a baseline drawn at random from a uniform distribution, $\mathcal{U}(-1,1)$ all datasets. This aligns with standard evaluation practices for IG.

@@

As IG outputs are not bounded (i.e., they are arbitrary real numbers), it becomes a challenge to meaningfully compare IG outputs of different models — ones that are trained conventionally, and ones that underwent counterfactual training. For our purposes, we observe that the direction of contribution (sign) does not matter. If protecting a feature flips the sign of the corresponding output, then CT „overcorrected” for this feature, but as long as the magnitude of its contribution decreases, then CT achieved its objective according to our formulation. Thus, we focus on the magnitude of IG outputs in the evaluation. Next, to accommodate for variable ranges of outputs, we standardize the outputs.

@@

::: {layout-ncol=2}

@@

![Class-conditional standardized integrated gradients for MNIST. Top: BL. Bottom: CT. Mutability constraints were imposed on the five top and bottom rows of pixels.](/paper/figures/mnist_ig.png){#fig-mnist-ig}

>>>>

To calculate the contributions, IG compares the output to a baseline. The selection of an appropriate baseline is an important design decision — it should produce a "neutral" prediction to avoid capturing effects that cannot be directly attributed to the model [@sundararajan2017ig; @sturmfels2020visualizing]. To remain consistent in our evaluations, we use a baseline drawn at random from a uniform distribution, $\mathcal{U}(-1,1)$, for all datasets. This aligns with standard evaluation practices for IG.

@@

As IG outputs are not bounded (i.e., they are arbitrary real numbers), it becomes a challenge to meaningfully compare IG outputs of different models — ones that are trained conventionally, and ones that underwent counterfactual training. For our purposes, we observe with reference to our Proposition, that we are interested estimating changes in the relative contribution of protected features compared to mutable ones. Thus, to meaningfully compare integrated gradients for different models and to accommodate for variable ranges of outputs in absolute terms, we standardize the integrated gradients across features.

@@

@@
